INLA replication

### 2022-09-15 src/example_inla-replication/inla-replication.Rmd
Upload crazed attempts from two years ago to do this -- likely very confused but perhaps useful things in there to unpick now

---
title: "Blangiardo replication"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    **Background**

    **Task**

    **Findings**

    **Next steps**
---

Step-by-step example of the INLA method.
The model for i.i.d. observations $y = (y_1, \ldots, y_n)$ is as follows
\begin{align}
y_i \, | \, x, \theta &\sim \mathcal{N}(x, 1/\theta), \\
x &\sim \mathcal{N}(x_0 = -3, 1/\tau_0 = 4), \\
\theta &\sim \text{Gamma}(a = 1.6, b = 0.4).
\end{align}

## Set-up

```{r message=FALSE, warning=FALSE}
midblue <- "#3D9BD0"
midgreen <- "#00855A"
midpink <- "#B3608E"
```

Load the data:

@@

```{r, fig.height=2, fig.cap="The observations (ordered by index, though this has no interpretation). The prior (mean shown as dashed line) mismatches with the data."}
data.frame(index = 1:30, y = y) %>%
  ggplot(aes(x = index, y = y)) +
  geom_point(shape = 4) +
  labs(title = "The data") +
  geom_hline(yintercept = -3, col = "#666666", lty = "dashed") +
  theme_minimal()
```

Parameters of the prior distribution:

@@

Exact distribution of $x$ given $\theta$:

```{r}
inner_loop <- function(theta) {
  tau_n <- n * theta + tau_0
  x_n <- (theta * n * y_bar + tau_0 * x_0) / tau_n
  return(list(x_n = x_n, tau_n = tau_n))
}
```

@@

For INLA's grid strategy, starting from the mode, we take steps of size $\delta_z$ checking that each point meets the criteria
\begin{equation}
  \log \tilde p(\theta(0) \, | \, y) - \log \tilde p(\theta(z) \, | \, y) < \delta_\pi. \label{eq:criteria}
\end{equation}
In the one dimensional case there is no need to do the $z$-parametrisation.
Choosing $\delta_z$ and $\delta_\pi$ based upon manual tuning (how does INLA select these numbers in general?):

@@

ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  annotate("text", x = 0.175, y = 5e-41,
           label = "Now the points are concentrated \n where there is higher density")
```

One of the INLA vignettes discusses how the user can set their own integration points--see `browseVignettes(package = "INLA")`.
The integration points that INLA uses can be found using `fit$joint.hyper`, which we can add on to the above plot:

@@

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  annotate(
    "text", x = 0.175, y = 5e-41,
    label = "Now the points are concentrated \n where there is higher density"
  ) +
  theme_minimal()
```

@@

For any given input $x$ you can do quadrature according to
\begin{equation}
  \tilde p(x \, | \, y) =
  \sum_{k = 1}^K p(x \, | \, \theta^{(k)}, y) \times p(\theta^{(k)} \, | \, y) \times \Delta^{(k)},
\end{equation}
but how should the range of $x$ be chosen?
Blangiardo again use a naive grid for demonstration:

@@

The $x$ and $y$-axis do not correspond to values, just to indicies:

```{r}
image(nl_joint_post)
```

@@

```{r}
its <- 1000

@@

```{r}
#' For some reason the mode in the INLA result below is NA
kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$mean),
                           exp(r_opt$par),
                           exp(tmb_opt$par)))

@@

>>>>

---
title: "INLA replication"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    | We demonstrate a step-by-step example of the INLA method implemented in R and TMB.
---

```{r setup, message=FALSE, warning=FALSE, class.source = 'fold-hide'}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)

library(tidyverse)
library(TMB)
library(INLA)
library(rmarkdown)

midblue <- "#3D9BD0"
midgreen <- "#00855A"
midpink <- "#B3608E"
```

# Background

Consider a three-stage model
\begin{alignat}{2}
  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \,\mathbf{0},  \mathbf{Q}(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
\end{alignat}
with posterior
\begin{equation}
  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
\end{equation}

After writing the negative log-target in a C++ file `model.cpp` it can be compiled and loaded into R:

```{r eval=FALSE}
compile("model.cpp")
dyn.load(dynlib("model"))
```

Let $f(x, \theta) \equiv - \log p(y, x, \theta)$ where $y$ is assumed to be fixed.
This objective (`f$fn`) and its derivative (`f$gn`) can be created (sketch) in `TMB` using:

```{r eval=FALSE}
f <- MakeADFun(data = dat, parameters = param, DLL = "model")
```

There are four stages to replicating the INLA method.

## Stage 1: get $\tilde p(\theta \, | \, y)$

<!-- Taking negative logs (to incorporate $f$) gives -->
<!-- \begin{equation*} -->
<!-- - \log p(\btheta \, | \, \y) \sim f(x, \btheta) + \log p(x \, | \, \btheta, \y) -->
<!-- \end{equation*} -->

Approximate the posterior marginal $p(\theta \, | \, y) \propto L(\theta)$ using the Laplace approximation $L(\theta) \approx L^\star(\theta)$:
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu^\star(\theta)} \equiv L^\star(\theta)
\end{equation}

Using `TMB` if we integrate out the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:

```{r eval=FALSE}
h <- MakeADFun(data = dat, parameters = param, DLL = "model", random = "x")
```

Therefore, `exp(-h$fn(theta))` is proportional to $\tilde p(\theta \, | \, y)$:

<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\btheta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\btheta)) + f(\bm{\mu}^\star(\btheta), \btheta)$?* -->

## Stage 2: choose $\{\theta_k, \Delta_k\}_{k = 1}^K$

Integration points $\{\theta_k\}$ with weights $\sum_{k = 1}^K \Delta_k = 1$.

### Empirical Bayes

$K = 1$ with integration point $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
This can be found by minimising `h` using optimisers like `nlminb` or `optim`.

### INLA grid

The steps required are:

1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
2. Calculate negative Hessian at the mode.
The solution is likely to use some version of `h$env$spHess` although it's not clear which one.
3. Test points against the $\delta_z$ condition

### Adaptive Gaussian Hermite quadrature

## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$

### Gaussian

Obtain marginals of $\tilde p_G(x \, | \, \theta, y)$ from above.

### Laplace

Compute the Laplace approximation

### Simplified Laplace

## Stage 4: apply quadrature

\begin{equation}
  \tilde p(x_i \, | \, y) = 
  \sum_{k = 1}^K \tilde p(x_i \, | \, \theta^{(k)}, y) \times \tilde p(\theta^{(k)} \, | \, y) \times \Delta^{(k)}
\end{equation}

# Example 1: Blangiardo

Consider the following simple model for i.i.d. observations $y = (y_1, \ldots, y_n)$
\begin{align}
y_i \, | \, x, \theta &\sim \mathcal{N}(x, 1/\theta), \\
x &\sim \mathcal{N}(x_0 = -3, 1/\tau_0 = 4), \\
\theta &\sim \text{Gamma}(a = 1.6, b = 0.4).
\end{align}

We start by loading the data manually:

@@

The observations (ordered by index, though this has no interpretation) and prior (mean shown as dashed line, which is significantly different to the data):

```{r, fig.height=2, fig.cap=""}
data.frame(index = 1:30, y = y) %>%
  ggplot(aes(x = index, y = y)) +
  geom_point(shape = 4) +
  geom_hline(yintercept = -3, col = "#666666", lty = "dashed") +
  theme_minimal()
```

We use the following parameters of the prior distribution:

@@

The distribution $p(x \, | \, \theta, y)$ is exact:

```{r}
# Exact distribution of x given theta
inner_loop <- function(theta) {
  tau_n <- n * theta + tau_0
  x_n <- (theta * n * y_bar + tau_0 * x_0) / tau_n
  return(list(x_n = x_n, tau_n = tau_n))
}
```

@@

For INLA's grid strategy, starting from the mode, we take steps of size $\delta_z$ checking that each point meets the criteria
\begin{equation}
  \log \tilde p(\theta(0) \, | \, y) - \log \tilde p(\theta(z) \, | \, y) < \delta_\pi. \label{eq:criteria}
\end{equation}
In the one dimensional case there is no need to do the $z$-parametrisation.
Choosing $\delta_z$ and $\delta_\pi$ based upon manual tuning^[It would be good to know how INLA selects these numbers in general]:

@@

ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  annotate(
    "text", x = 0.175, y = 5e-41,
    label = "Now the points are concentrated \n where there is higher density"
  ) +
  theme_minimal()
```

One of the INLA vignettes discusses how the user can set their own integration points -- see `browseVignettes(package = "INLA")`.
The integration points that `R-INLA` uses can be found using `fit$joint.hyper`, which we can add on to the above plot for interest:

@@

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  theme_minimal()
```

@@

For any given input $x$ you can do quadrature according to
\begin{equation}
  \tilde p(x \, | \, y) =
  \sum_{k = 1}^K p(x \, | \, \theta^{(k)}, y) \times p(\theta^{(k)} \, | \, y) \times \Delta^{(k)},
\end{equation}
but how should the range of $x$ be chosen?
Blangiardo use a naive grid for demonstration:

@@

The $x$ and $y$-axis do not correspond to values here, just indices:

```{r}
reshape2::melt(nl_joint_post) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) + 
    geom_tile() +
    theme_minimal() +
    labs(fill = "", x = "x", y = "theta") +
    theme(legend.position = "none")
```

@@

```{r message=FALSE}
its <- 1000

@@

```{r}
# For some reason the mode in the INLA result below is NA, how to fix this?
kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$`0.5quant`),
                           exp(r_opt$par),
                           exp(tmb_opt$par)))

@@

### Is `h$fn` the same as in R?

Testing to see that `h$fn` is the same as `nl_post_marginal_theta` with `log_input = TRUE` up to a constant:

```{r results=FALSE}
vals <- seq(-5, 0, by = 1)

tmb_vals <- sapply(vals, h$fn) # Replications show it not to be stochastic
r_vals <- sapply(vals, nl_post_marginal_theta, log_input = TRUE)
```

```{r}
rbind(tmb_vals, r_vals, tmb_vals - r_vals)
```

Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:
```{r results=FALSE}
tmb_post_marginal_theta <- function(theta) {
  de_nl(h$fn, x = log(theta))
} # (Have tested that this function does as it should)

tmb_dense_theta <- eval_grid(
  uniform = TRUE,
  K = 500, min = 0.001, max = 0.3, 
  f = tmb_post_marginal_theta
)
```

```{r fig.cap = "At least this verifies that the optimisation is sensible."}
# Normalised dense_theta and tmb_dense_theta since they include different constants
ggplot(dense_theta) +
  geom_line(aes(x = input, y = norm_output), col = midblue) +
  geom_line(data = tmb_dense_theta, aes(x = input, y = norm_output), col = midpink) +
  theme(axis.text.y=element_blank()) +
  annotate("text", x = 0.1, y = 0.01, label = "TMB", col = midpink) +
  annotate("text", x = 0.045, y = 0.012, label = "R", col = midblue) +
  theme_minimal()
```

Does `TMB` do the inner optimisation differently?

```{r results=FALSE, message=FALSE}
get_tmb_inner_optimised_value <- function(l_theta) {
  sd_out <- sdreport(h, par.fixed = l_theta)
  reml <- sd_out$par.random # The REML estimator for x from TMB
  return(reml)
}

tmb_inner <- sapply(vals, get_tmb_inner_optimised_value)
r_inner <- sapply(exp(vals), function(theta) inner_loop(theta)$x_n)
```

```{r}
rbind(tmb_inner, r_inner)
```

Looks like this part is correct.

### Getting to the full conditional of $x$ in TMB

Probably going to need $Q^\star(\btheta)$ ($n \times n$ precision matrix) and $\bm{\mu}^\star(\btheta)$ ($n \times 1$ mean vector).
For example, the most basic version of the full conditional would be the Gaussian approximation
\begin{equation}
\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\btheta), 1 / q^\star_i(\btheta)).
\end{equation}

Precision matrix $Q^\star(\btheta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\bm{\mu}^\star(\btheta)$, having $(i, j)$th entry
\begin{equation}
Q^\star(\btheta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \btheta, \y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \bm{\mu}^\star(\btheta)}
\end{equation}

An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects".
There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way.
This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format.

`h$env$spHess` takes two arguments:

  1. `par` (default `h$env$par`) should be both the hyperparameters and random effects
  2. `random` (`TRUE` or `FALSE`)

and the code is given by:

```{r eval=FALSE}
function(par = obj$env$par, random = FALSE) {
  if (!random) {
    Hfull@x[] <- ev(par) # Whats ev?
    Hfull # H full <=> Full Hessian?
  }
  else if (skipFixedEffects) {
    .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB")
  }
  else {
    Hfull@x[] <- ev(par)
    Hfull[r, r]
  }
}
```

Set a range of test parameters `test_pars`:

```{r results=FALSE}
full_pars <- function(input) {
  setNames(input, c("x", "l_theta"))
}

par1 <- full_pars(c(0, 0))
par2 <- full_pars(c(0, -1))
par3 <- full_pars(c(-1, 0))

# Doing the inner loop manually
par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0))
par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1))

test_pars <- list(par1, par2, par3, par4, par5)
```

```{r}
lapply(test_pars, h$env$spHess)
```

`random = TRUE` seems behave in unpredictable ways e.g. always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted):

```{r results=FALSE}
lapply(test_pars, h$env$spHess, random = TRUE)
```

1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters).
The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$.
2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix.

Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters:

```{r}
lapply(test_pars, f$he)
```

This produces the same answers as numerical differentiation of the same function (results omitted for space):

```{r results=FALSE}
lapply(test_pars, optimHess, fn = f$fn, gn = f$gn)
```

In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!):

```{r}
sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n)
```

Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R).

```{r}
nl_full_conditional_x <- function(x, theta, log_input = FALSE) {

  if(log_input == TRUE) {
    theta <- exp(theta) # Convert from log(theta) to theta
  }

  par <- inner_loop(theta)

  return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE))
}
```

# Example 2: Unknown
